---
title: "A Gentle Introduction to Bayesian Statistics"
output: 
  github_document:
    fig_width: 6
    fig_height: 4
---

This repository will contain course material for the one-day course "A Gentle Introduction to Bayesian Statistics" at the [Epinor Annual Meeting 2021](https://en.uit.no/forskning/forskningsgrupper/nyheter/artikkel?p_document_id=476300&sub_id=744901). 

The tentative plan for the day is as follows. Links to the course material will be added once it's ready.

| Time | What |
| --- | --- |
| 10:00 - 11:30 | Part I - Introduction to Bayesian Statistics |
| 11:30 - 12:30 | Lunch |
| 12:30 - 14:00 | Part II - Bayesian Workflow for Data Analysis |
| 14:15 - 15:15 | Part III - Bayesian Analysis with R |
| 15:30 - 17:00 | Exercises with "brms" in R |

The course material is not quite ready yet, but below are some instructions for those who want to have the 'brms' packages properly installed before we start.

# Setup

If you want to follow the last session, "Exercises with 'brms' in R", please follow the instructions here to make sure your system is properly set up. Installing 'brms' is slightly more complicated than installing most other R packages, and I will of course also help with this on-site.

First, if you don't have it already, [download RStudio](https://www.rstudio.com/products/rstudio/download/#download).

Next, make sure you have R version 4.1.0 or later installed. If in doubt, open RStudio and type `sessionInfo()`. It should give you something like this, with the R version printed on the first line.

```{r}
sessionInfo()
```

If the version is too old, [go here to download the latest version](https://cloud.r-project.org/).

Next, we need to install 'brms'. The instructions below are hopefully sufficient, but if you have trouble, please write me an e-mail or (even better) open an [Issue in this repository](https://github.com/osorensen/BayesCourseEPINOR2021/issues), since then everyone else will be able to see.

## Mac OS

If you're on a Mac, make sure you have Xcode installed. You can install Xcode by [clicking this link which you can open in App Store](https://apps.apple.com/us/app/xcode/id497799835?mt=12). 

After you've installed Xcode, issue this command:

```{r, eval=FALSE}
install.packages("brms")
```

## Windows

On Windows, you first need to install RTools. [Go here, and follow the installation instructions](https://cran.r-project.org/bin/windows/Rtools/). You're very likely on Windows 64-bit.

After you've installed RTools, restart the computer, open RStudio and type:

```{r, eval=FALSE}
install.packages("brms")
```

